### 2016-07-01 image_setup.tex
Image resizing edits

The latest Raspbian lite bundle can be downloaded from the Raspbian
official webpage~\cite{raspbian}. At the time of this writing, the latest
available bundle was \texttt{2016-05-27-raspbian-jessie-lite.zip}.
To ensure that the content of the bundle does not change, this procedure
is based on that particular version of Raspbian lite which we have
made available at~\cite{tunescode_webpage}. All other files used in this paper
are available there unless specified otherwise. To get started, open a
Linux shell (terminal) and declare the environment variables.
To copy the commands from this procedure to the testbed administrator terminal,
we strongly recommend to read this document with Adobe Acrobat Reader in Linux.
Otherwise, copied characters might not render properly.

@@

In the code block, the \texttt{URL} and \texttt{IMAGE} specify where
the Linux bundle is located and \texttt{WORKDIR} specifies a working
directory where the Raspbian bundle will be downloaded and customized.
Notice that even though we use the \$ and \# signs in the shell, in general
these signs will be particular to the testbed administrator \ac{OS} shell.
Next, create the working directory and go to it with the \texttt{cd} command.
To download the image, utilize \texttt{wget} before
unpacking the \texttt{zip} file as follows:

@@

After Raspbian lite has been unpacked, there should be an \texttt{img} file in the directory.
\texttt{fdisk} can be used to display the content of the image. Parse the arguments
"-u sectors" to display the sizes in sectors and "-l" to display the partitions within the
image and and automatically exit \texttt{fdisk} again:

@@

The output provides knowledge about the image. The image is in total
2709504 sectors (1.3GiB) in size and contains two
partitions. The first partition starts at sector 8192 and
the other partition starts at sector 137216. The partitions is of type FAT32
and Linux and are 63M and 1.2G big. This indicates that the first partition
is most likely a boot partition and the second partition is a traditional Linux file
system. Actually, it is the Linux root filesystem (i.e. /).

\subsection{Resize image}
1.2G is not very much to store the root filesystem given that it is desired to
alter the files and install additional packages. It may therefore be
necessary to increase the partition size.
The following illustrates how the image and its root filesystem
can be expanded by one \ac{GB}.

First, expand the image one \ac{GB} and wait for the data to be completely written
to the disk:
\begin{lstlisting}[]
$ dd if=/dev/zero bs=1M count=1024 >> ${IMAGE}.img && sync
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

Use \texttt{fdisk} to see that the image is now one \ac{GB} larger:
\begin{lstlisting}[literate={DOLLAR}{\$}1]
DOLLAR fdisk -u sectors -l ${IMAGE}.img
Disk 2016-05-27-raspbian-jessie-lite.img: 2.3 GiB, 2461007872 bytes, 4806656 sectors
Units: sectors of 1 * 512 = 512 bytes
Sector size (logical/physical): 512 bytes / 512 bytes
I/O size (minimum/optimal): 512 bytes / 512 bytes
Disklabel type: dos
Disk identifier: 0x6fcf21f3

@@

Now, to expand the root filesystem it is required to first remove the Linux partition
and then add it again one \ac{GB} bigger.
Use \texttt{fdisk} to alter the partition table. The \texttt{echo}'s within the parenthesis
will be interpreted as
key-presses in \texttt{fdisk} and hence delete the last partition, create a new
primary partition, print new partition table, and write new partition table
to the image:

@@

Use a loopback device to make the Raspbian image available as a block device: 
\begin{lstlisting}[]
# export DEV=$(sudo losetup --show -f -P ${IMAGE}.img); echo $DEV
/dev/loop0
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

>>>>

The latest Raspbian lite bundle can be downloaded from the Raspbian
official webpage~\cite{raspbian}. At the time of this writing, the latest
available bundle was \texttt{2016-05-27-raspbian-jessie-lite.zip}.
To ensure that the content of the bundle does not change, this procedure
is based on that particular version of Raspbian lite which we have
made available at~\cite{tunescode_webpage}. All other files used in this paper
are available there unless specified otherwise. To get started, the testbed
administrator must open a Linux shell (terminal) in and declare the environment
variables shown in the command block below. We show the procedure by
performing the role of the testbed administrator. To copy the commands from
this procedure to the testbed administrator terminal,
we strongly recommend to read this document with Adobe Acrobat Reader in Linux.
Otherwise, copied characters might not render properly.

@@

In this code block, the \texttt{URL} and \texttt{IMAGE} variables specify where
the Linux bundle is located and \texttt{WORKDIR} specifies a working
directory where the Raspbian bundle will be downloaded and customized.
Notice that even though we use the \$ and \# signs in the shell, in general
these signs will be particular to the testbed administrator \ac{OS} shell.
Next, we create the working directory and go to it with the \texttt{cd}
command. To download the image, we utilize the \texttt{wget} command before
unpacking the \texttt{zip} file as follows:

@@

After Raspbian lite has been unpacked, there should be an \texttt{.img}
file in the \texttt{WORKDIR} directory. Here, \texttt{fdisk} can be used to
display the content of the image. To do this and obtain administrative
information, we parse the arguments \texttt{-u sectors} to display the
sizes in sectors and \texttt{-l} to display the partitions within the
image. The \texttt{fdisk} command should output to the terminal something
similar to:

@@

The output provides relevant information about the image. The image is in
total 2709504 sectors (1.3GiB) in size and contains two partitions. The
first partition starts at sector 8192 and the other partition starts at
sector 137216. The first partition type is FAT32 with a size of 63 MB
and the second partition is a Linux one with a size of 1.2 GB. This
indicates that the first partition is a boot partition and the second
one is a traditional Linux root filesystem, i.e. \texttt{/}.

\subsection{Image Resizing}
Given that we want to customize the files stored in the \ac{Raspi}s,
we need to resize the image file since 1.2GB might not be enough to store
the root filesystem due to the total size of the additional packages. Thus,
we need to increase the partition size. The following procedure illustrates
how the image and its root filesystem can be expanded by one \ac{GB}.
First, to expand the image one \ac{GB}, we execute:

\begin{lstlisting}[]
$ dd if=/dev/zero bs=1M count=1024 >> ${IMAGE}.img && sync
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

Later, we use \texttt{fdisk} with the same arguments as before to see that
the image is now one \ac{GB} larger:
\begin{lstlisting}[literate={DOLLAR}{\$}1]
DOLLAR fdisk -u sectors -l ${IMAGE}.img
Disk 2016-05-27-raspbian-jessie-lite.img: 2.3 GiB, 2461007872 bytes, 4806656 sectors
Units: sectors of 1 * 512 = 512 bytes
Sector size (logical/physical): 512 bytes / 512 bytes
I/O size (minimum/optimal): 512 bytes / 512 bytes
Disklabel type: dos
Disk identifier: 0x6fcf21f3

@@

Now, in this command block output, we observe that the change has taken
effect by noticing the total available image size is 2.3GiB. To expand the
root filesystem, it is required to first remove the Linux partition
and then add it again with one \ac{GB} more. To do this, we make use
again of \texttt{fdisk} in command mode to alter the partition table.
For this, we pass the arguments of the command mode of \texttt{fdisk}
through the \texttt{echo} command in Linux and the \texttt{|} operator
(pipe) as follows:

@@

The \texttt{echo} commands within the parenthesis are interpreted as
key-presses in the \texttt{fdisk} commmand mode. They (i) delete the last
partition, (ii) create a new primary partition, (iii) print new partition
table and (iv) write new partition table to the image file. Use a loopback
device to make the Raspbian image available as a block device:

\begin{lstlisting}[]
# export DEV=$(sudo losetup --show -f -P ${IMAGE}.img); echo $DEV
/dev/loop0
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}
